MS_Intro_Short

### 2023-10-14 Markdown/Manuscript/Child_RMDs/MS_Aims.Rmd
redo of intro

In this paper, we describe:

(i)	criteria for assessing ethical CHEM implementations;

(ii) a prototype software framework for the ethical implementation of CHEMs; and

(iii)	use of the software framework to develop an open-source CHEM in youth mental health, with an initial focus on outcome valuation.

>>>>

In this paper, we:

(i)	propose core ethical responsibilities for CHEM developers and criteria for assessing ethical CHEM implementations;

(ii) describe a prototype software framework we have developed to support the ethical implementation of CHEMs; and

(iii)	assess an initial set of youth mental health CHEMs developed with our software framework against our proposed ethical CHEM implementation criteria.

### 2023-10-14 Markdown/Manuscript/Child_RMDs/MS_Intro_Short.Rmd
redo of intro

Health economics is a discipline concerned with problems that arise due to scarce resources, such as how to value health and healthcare, allocate healthcare budgets and configure health services [@wagstaff2012four]. In seeking to solve these problems, health economists typically use models which are simplified and selective representations of systems that are believed to influence human health. A health economic model should be capable of being described using words and figures (a conceptual model), equations (a mathematical model) and software (a computational model).  Health economic scientific manuscripts typically describe a model using its conceptual and mathematical representations but report results that have been generated by its computational representation (i.e., through execution of a computer program that applies the computational model to selected input data). The conceptual, mathematical and computational representations of a model are assumed to be isomorphic. However, the ease of independently assessing the validity of this assumption depends in part on how a health economic model has been implemented as software.

<!-- the execution of a computer program that applies a computational model to compatible data inputs (e.g., parameter values) and performs a sequence of numeric calculations. -->
<!-- the predictions reported in health economic studies are now almost always -->

Computational health economic models (CHEMs) can be implemented using specialised commercial modelling software or authored as software development projects in an open-source programming language such as Python [@python2009] or R [@RCORE2022]. Advantages of commercial modelling tools include simplicity and ease of use, but an open-source software development approach may facilitate computational models that are more transparent, reusable and updatable [@incerti2019r; @Pouwels2022]. The decision about what implementation approach to adopt needs to balance both modelling team value judgments about priority model features and the resource and time constraints within which a CHEM project needs to be delivered (see Table \@ref(tab:proscons)).

```{r, child = child_docs_ls$one, echo=FALSE}
```
<!-- on considerations such as model comprehensiveness, generalisability, reproducibility and transferability -->
For open-source software projects, a major consideration is the selection of an appropriate software framework. A software framework is a shared common technology used by developers to collaboratively author software and which is not typically visible to software end-users [@myllarniemi2018development]. A software framework provides a foundation for developing multiple software applications with shared resources (e.g., code and data files), that can be modified to suit specific needs. Software frameworks have widely been developed and implemented in data science, for example, PyTorch [@PyTorch2019] for machine learning using Python.

Advantages of using software frameworks include facilitating code reuse and extension, promoting good programming practice and the capability to provide enhanced functionality and performance without additional effort by developers [@edwin2014software]. However, software frameworks can be challenging and time consuming to create [@edwin2014software] and then difficult for others to learn, often requiring model developers to undergo specialist training [@myllarniemi2018development]. There is also a risk that a software framework may become excessively complex over time [@edwin2014software].

<!-- We are developing a model to explore multiple economic questions relating to the mental health of young people aged 12 to 25.   -->
<!-- When making choices about how to implement this health economic model computationally, we wished to facilitate ethical development and use. -->
The ethical responsibilities of computational modellers seeking to inform policy decisions are often poorly understood and inadequately fulfilled [@10.3389/fpubh.2017.00068; @pliakos2021ethics; @thompson2022escape; @calder2018computational]. Within the health economics discipline, open-source CHEMs remain rare [@Feenstra2022; @Emerson2019]. The main barriers to open-source CHEMs identified in a 2020 survey of health economists concern challenges relating to updating models, legal considerations, transferring model data, the platforms or languages in which models are developed, the level of public access required, lack of interest from decision-makers, model generalisability and confidentiality and security considerations [@Pouwels2022].

There is a potential role for software frameworks to help health economic modellers overcome some of these barriers. A high-level software framework for implementing open-source CHEMs in R has previously been developed with the primary aim of improving model transparency [@alarid2019need]. Ideally, software frameworks would facilitate fulfillment of a broader set of ethical responsibilities. However, guidance on developing, selecting and using software frameworks for ethical CHEM implementations is scarce.

>>>>

Health economics is a discipline concerned with problems that arise due to scarce resources, such as how to value health and healthcare, allocate healthcare budgets and configure health services [@wagstaff2012four]. In seeking to solve these problems, health economists typically use models which are simplified and selective representations of systems that are believed to influence human health. A health economic model should be capable of being described using words and figures (a conceptual model), equations (a mathematical model) and software (a computational model).  Health economic scientific manuscripts typically describe a model using its conceptual and mathematical representations but report results that have been generated by its computational representation (i.e., in a format that allows computers to perform calculations). The conceptual, mathematical and computational representations of a model are assumed to be isomorphic. However, the ease of independently assessing the validity of this assumption depends in part on how a health economic model has been implemented computationally.
<!-- through execution of a computer program that applies the computational model to selected input data). -->
<!-- the execution of a computer program that applies a computational model to compatible data inputs (e.g., parameter values) and performs a sequence of numeric calculations. -->
<!-- the predictions reported in health economic studies are now almost always -->

The set of digital files implementing a health economic model's structure, data and algorithms as software can be called a computational health economic model (CHEM). CHEMs can be developed using specialised commercial modelling software or in an open-source programming language such as Python [@python2009] or R [@RCORE2022]. An advantage of commercial modelling software is that users often require no or limited programming skills to develop and apply CHEMs. However, open-source software programming languages may facilitate the development of CHEMs that are more transparent, reusable and updatable [@incerti2019r; @Pouwels2022]. The decision about what software development approach to adopt needs to balance both modelling team value judgments about priority CHEM features and the skills, resource and time constraints within which a CHEM project needs to be delivered (see Table \@ref(tab:proscons)).

```{r, child = child_docs_ls$one, echo=FALSE}
```

Within the health economics discipline, open-source CHEMs remain rare [@Feenstra2022; @Emerson2019]. The main barriers to open-source CHEMs identified in a 2020 survey of health economists concern challenges relating to updating models, legal considerations, transferring model data, the platforms or languages in which models are developed, the level of public access required, lack of interest from decision-makers, model generalisability and confidentiality and security considerations [@Pouwels2022]. 
<!-- A major consideration in open-source software development, is the selection of an appropriate software framework.  -->
<!-- on considerations such as model comprehensiveness, generalisability, reproducibility and transferability -->

Some of these barriers could potentially be mitigated through the development of appropriate software frameworks. A software framework is a shared common technology used by developers to collaboratively author software and which is not typically visible to software end-users [@myllarniemi2018development]. A software framework provides a foundation for developing multiple software applications with shared resources (e.g., code and data files), that can be modified to suit specific needs. Advantages of using software frameworks include facilitating code reuse and extension, promoting good programming practice and the capability to provide enhanced functionality and performance without additional effort by developers [@edwin2014software]. Software frameworks have widely been developed and implemented in data science, for example, PyTorch [@PyTorch2019] for machine learning using Python.

<!-- We are developing a model to explore multiple economic questions relating to the mental health of young people aged 12 to 25.   -->
<!-- When making choices about how to implement this health economic model computationally, we wished to facilitate ethical development and use. -->

A high-level software framework for implementing open-source CHEMs in R has previously been developed with the primary aim of improving model transparency [@alarid2019need]. A software framework that help CHEM developers fulfill a broader set of ethical responsibilities would require a number of conceptual, technical and behavioural challenges to be overcome. Firstly, the ethical responsibilities of developers of computational models designed to inform governmental decision-making are often poorly understood <!-- and inadequately fulfilled  -->[@10.3389/fpubh.2017.00068; @pliakos2021ethics; @thompson2022escape; @calder2018computational]. Software frameworks can be challenging and time consuming to create and may become excessively complex over time [@edwin2014software]. Furthermore, software frameworks can be difficult to learn, often requiring model developers to undergo specialist training [@myllarniemi2018development].
<!-- a software framework is unlikely to be widely adopted unless the benefits to developers outweigh the costs associated with using it. -->
<!-- However, guidance on developing, selecting and using software frameworks for ethical CHEM implementations is scarce. -->
